---
title: "Statistical analysis (main dependent variable)"
output:
  html_document:
    df_print: paged
---
<font size="4"> Brent J. Raiteri (on data from Paolo Tecchio's hybrid 3D fascicle tractography approach) <br>
2026-02-27 </font>

## Fascicle length from DTI
Fixed effects were fitted using a linear mixed-effects model. Model-based standard errors and p-values are presented. Sandwich variance estimation-based standard errors and p-values were also produced; these do not assume normality of the data.

```{r setup_global, include=FALSE}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, digits = 4)

```

```{r setup_libraries,include=FALSE}

# Libraries
library(tidyverse) #also loads tibble and ggplot2
library(lme4)
library(flexplot)
library(parameters)
library(clubSandwich)
library(lmerTest)
library(emmeans) # posthoc tests
```
```{r get_data,include=FALSE}
# Get data
d = read.csv('C:\\Users\\Brent\\Desktop\\Research\\PhD\\Paolo\\tecchioNature.csv')

# Specify factors
d <- d %>%
  mutate(
    id      = as.factor(id),
    angle   = as.factor(angle),
    leg = as.factor(leg),
    volus = as.numeric(volus),
    voldti = as.numeric(voldti),
    lenus = as.numeric(lenus),
    lendti = as.numeric(lendti)
  )

```

```{r boxplot, echo=FALSE}

## Graphical display
ggplot(d, aes(x = angle, y = lendti, fill = leg)) +
       geom_boxplot() +
       #facet_wrap(~time, labeller = as_labeller(c("1" = "PRE", "2" = #"POST"))) +  # Adds the 4th variable (time) as separate panels
       labs(x = "Plantar flexion angle (deg)",
                       y = "Fiber length (mm)",
                       fill = "Leg", # Label the legend
                       title = "Graphical display") + 
  scale_fill_discrete(labels = c("Left", "Right")) + 
  # Use scale_x_discrete() to define the new labels for the x-axis
  scale_x_discrete(
    labels = c(
      "1" = expression("15"),
      "2" = expression("35")
    )
  )

```

## Fixed-effects model with interaction 
This fixed-effects model fits the factors plantar flexion angle and leg with random intercepts and fixed slopes and a fixed angle x leg interaction. <br> 
<br>
We first compared fitting the model by REML vs. ML, and REML resulted in a better fit:
```{r fit_interaction_model, result="hide", echo=FALSE}

interaction_REML <- lmer(lendti ~ angle * leg + (1|id), data = d, REML=TRUE)

interaction_ML <- lmer(lendti ~ angle * leg + (1|id), data = d, REML=FALSE)

model.comparison(interaction_REML,interaction_ML)

#interaction_TrT <- lmer(torque ~ training * time + condition + (1|id), data = d, REML=TRUE)

#model.comparison(interaction_REML,interaction_TrT)

```
## Model summary
```{r summary_interaction_model, echo=FALSE}
summary(interaction_REML)
```
## Model- & sandwich-based standard errors and p-values
The full cluster-robust variance-covariance matrix was calculated, which includes the variances (SEs) and covariances (how much the estimates correlate with each other):
```{r comparison_interaction_model,echo=FALSE}

# Assuming 'interaction_REML' is your fitted lmer model object
# --- 1. Extract Model-Based Results using the parameters package ---
model_results_raw <- model_parameters(interaction_REML) %>%
  as_tibble() %>%
  filter(!str_detect(Parameter, "^SD \\("))

# Use the correct names found: "Coefficient" and "SE"
model_results <- model_results_raw %>%
  select(term = Parameter, 
         estimate = Coefficient, # Corrected column name
         `Model-based s.e.` = SE,         # Corrected column name
         `Model-based p-value` = p)
  
# --- 2. Calculate Sandwich (Robust) Standard Errors ---
sandwich_vcov = vcovCR(interaction_REML, type="CR2")

# --- 3. Calculate Sandwich P-values and SEs using coef_test ---
# coeftest automatically uses the model's degrees of freedom
sandwich_test_results <- coef_test(interaction_REML, vcov = sandwich_vcov)

# Select the columns by their position: 
# Column 3 is 'SE', Column 7 is 'p-val (Satt)', and Column 2 is the term name ('beta')
sandwich_results <- as_tibble(sandwich_test_results) %>%
  select(term = 2,              # Selects the 2nd column for the term names
         `Sandwich s.e.` = 3,   # Selects the 3rd column for SE
         `Sandwich p-value` = 7) # Selects the 7th column for p-value

# Ensure the 'term' column is replaced by 'estimate'
sandwich_results <- rename(sandwich_results, estimate = term)

# --- 4. Combine into a Single Table ---
final_table <- model_results %>%
  left_join(sandwich_results, by = "estimate") %>%
  select(term, estimate, `Model-based s.e.`, `Model-based p-value`, `Sandwich s.e.`, `Sandwich p-value`)

# --- 5. Display the Table ---
cat("Model- & sandwich-based standard errors and p-values\n")
cat("Click on the arrow for sandwich-based p-values.\n\n")

# Print with Inf width/n for full display in console/notebook
print(final_table, n = Inf, width = Inf, digits=4)

```

## Estimated marginal means
Estimated marginal means (EMMs) for the fixed-effects model with a fixed angle x leg interaction are reported below, which explicitly use cluster-robust standard errors (accounting for non-independent data or heteroskedasticity). This approach was used to ensure that the confidence intervals and subsequent p-values for the EMMs are more accurate and reliable. <br>

```{r posthoc_comparisons,echo=FALSE}

# Generate EMMs with nesting
ems <- emmeans(interaction_REML, ~ angle | leg, 
                   vcov = sandwich_vcov, 
                   lmer.df = "satterthwaite")
print(ems)

avgLeg <- emmeans(ems, ~ angle)
print(avgLeg)

avgLeg_comparison <- pairs(avgLeg, reverse = TRUE)
print(avgLeg_comparison, adjust = "holm")

#indLeg <- pairs(ems)
#print(indLeg, adjust = "holm")

```